% The Clam manual
\def\rcsid{$Id: manual.tex,v 1.30 1999/04/30 14:42:32 img Exp $}
\input header
\expandafter\gdef\csname dofooter\endcsname{}

\thispagestyle{empty}
\pagestyle{empty}

\input   {Titlepage}
\subsection* {Abstract}
This note describes the \clam\ proof-planning system. It is intended
as a user manual for people who want to use \clam\ without knowing too
much about the insides, and as a programmer manual for people who want
to change and improve \clam. For quick reference, the note provides
appendices which summarise the most frequently used predicates in
\clam.


If you are only an interested reader and do not intend to actually use
\clam\ on a machine, then you should read only sections
\reference{intro} (Introduction) and \reference{methods} (The
Methods). This will give you a general idea of the capabilities of
\clam.

If you are a novice user and want to start playing straight away
without ploughing through 50 pages of manual, read section
\reference{intro} (the introduction), and section \reference{library} on
the library mechanism.


\paragraph {Acknowledgements.}  \clam\ is the result of a
collaborative effort between a number of members of the DReaM
group. The main ideas originated from Alan Bundy and were first
implemented by Frank van Harmelen;  Frank also wrote the first version
of this manual.  Alan Smaill contributed through
many suggestions and discussions, and wrote some of the methods and
tactics, Jane Hesketh wrote some of the early tactics, and Geraint
Wiggins was adventurous enough to be one of the first users.   Andrew
Stevens, Andrew Ireland and Ian Green developed \clam over the years.
Andrew and Ian made extensive contributions to this manual.

In addition to the above developers, the following people have made
contributions to the current \clam release: David Basin, Richard
Boulton, Jason Gallagher, Predrag Jani\v{c}i\'c, Helen Lowe, and
Santiago Negrete.  Other users and contributors include Francisco
Cant\'u Ortiz, Ina Kraan, Ra\'ul Monroy and Julian Richardson.  This
work is supported in part by {\sc epsrc} grants {\sc gr/h/23610}, {\sc
gr/j/80702} and {\sc gr/m/45030}.

\cleardoublepage

\pagestyle{fancy}
\pagenumbering{roman}
\tableofcontents

\include {Introduction}
\include {UserManual}
\include {ProgrammerManual}

\appendix
\include {general}
\include {appendix}

\lhead[\fancyplain{}{\sl BIBLIOGRAPHY}]{}
\rhead[]{\fancyplain{}{\sl BIBLIOGRAPHY}}
\bibliography{clam}
\bibliographystyle{plain}

\clearpage
\lhead[\fancyplain{}{\sl INDEX}]{\fancyplain{}{\sl Index of methods
and methodicals}}
\rhead[\fancyplain{}{\sl Index of methods}]{\fancyplain{}{\sl INDEX}}
\printindex[mthd]
\lhead[\fancyplain{}{\sl INDEX}]{\fancyplain{}{\sl Index of predicates}}
\rhead[\fancyplain{}{\sl Index of predicates}]{\fancyplain{}{\sl INDEX}}
\printindex[pred]
\clearpage
\lhead[\fancyplain{}{\sl INDEX}]{\fancyplain{}{\sl Index of Prolog
source files}}
\rhead[\fancyplain{}{\sl Index of Prolog source files}]{\fancyplain{}{\sl INDEX}}
\printindex[file]
\clearpage
\lhead[\fancyplain{}{\sl INDEX}]{}
\rhead[]{\fancyplain{}{\sl INDEX}}
\printindex[default]

\end{document}
